Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-08 paper/paper.tex
finished rewrtiting cplx-lrt

%notation
We first clarify some notation: $M^{\top}$ is the matrix transpose, $\conj{M}$ is elementwise
complex conjugation, and $M^{\hop} = (\conj{M})^\top$ denotes Hermitian conjugate. By $e_i$
we denote the $i$-th real unit vector of dimensionality \textbf{conforming} to the matrix-vector
expression it is used in. $\vec{M}$ denotes row-major flattening of $M$ into a vector, i.e.
in lexicographic order of its indices. $\otimes$ denotes the Kronecker product, for which we
note the following identities $
  \vec{P Q R} = (P \otimes R^\top) \vec{Q}
$, and $
  (P \otimes R) (C \otimes S) = P Q \otimes R S
$, \cite{petersen_matrix_2012}.
% It follows from the definition of the Kronecker product
% of $P$ and $R^\top$ and the row-major order vectorization.

@@

$, respectively. $\cplx$-Gaussians are closed under affine transformations: if $y = A z + b$
for $A \in \cplx^{n \times m}$ and $b \in \cplx^{n}$, then $
  y \sim \mathcal{N}_n^{\cplx}\bigl(

@@

$. The entropy of a $\cplx$-Gaussian distribution coincides with the entropy of

@@

$C = 0$. From \eqref{eq:c-gauss-entropy-derivation} the entropy of a proper $\cplx$-Gaussain is

@@

In $\cplx$-variational dropout we use mean-field proper $\cplx$-Gaussian variational
approximation. Specifically, parameters $W \in \cplx^{n\times m}$ of a dense layer follow
\begin{equation}  \label{eq:c-gauss-vi}
  % (we omit the subscript $m=1$ for brevity)
  W \sim q(W) = \prod_{ij}
    \mathcal{N}^{\cplx}(W_{ij} \mid \mu_{ij}, \sigma^2_{ij}, 0)
  \,,
\end{equation}
and we use analogous independent proper $\cplx$-Gaussian approximation for convolutional
kernels. As for bias terms, we propose to use their MAP estimates, i.e. assume an independent
fully factorized Dirac approximation.

As was demonstrated in \cite{kingma_variational_2015}, local reparameterization trick
provides both faster computations and better statistical efficiency of the SGVB gradient
estimator. We formulate the $\cplx$ version of the local reparameterization trick, which
generalizes \cite{wang_fast_2013}.

@@

To simplify exposition of the trick, we consider the complex linear dense layer, and afterwards
outline the caveats for $\cplx$-convolutional layers. In both cases we use fully factorized
general $\cplx$-Gaussian distribution for parameters $W$ of the layer.

Consider a $\cplx$-linear layer: $
  W \in \cplx^{n\times m}
$ is a random $\cplx$-Gaussian matrix with independent elements $
  w_{ij} \sim \mathcal{N}^{\cplx}\bigl(
    w_{ij}
    \big \vert
    \mu_{ij}, \sigma^2_{ij}, C_{ij}
  \bigr)
$ with $\mu_{ij}\in cplx$, $\sigma^2_{ij} \geq 0$, $C_{ij}\in \cplx$ and $
  \lvert C_{ij} \rvert^2 \leq \sigma^2_{ij}
$. We begin by observing that the variational approximation of $q(W)$ is a multivariate
$\cplx$-Gaussian: $
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{[n\times m]} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
$, where $[n\times m]$ denotes lexicographically flattened dimension and $\diag{v}$ is a
diagonal square matrix with $v_i$ on the main diagonal. Here we collected variances and means
into matrices $\Sigma$ and $\mu$, respectively.
%
Now, for any $x \in \cplx^m$ and $b\in \cplx^n$ we have
$$
W x + b
  % = I_n W x + b
  = (I_n \otimes x^\top) \vec{W} + b
  \,. $$
Structure of the matrices, involved in this expression, imply
\begin{align*}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^\hop
    % &= \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \sigma^2_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^\hop
    % \\
    &
    = \sum_{ij} (e_i \otimes x^\top e_j)
      \sigma^2_{ij} (e_i \otimes x^\top e_j)^\hop
    \notag \\
    % &= \sum_{ij} (e_i \otimes x_j) \sigma^2_{ij} (e_i^\top \otimes \conj{x}_j)
    % \\
    &
    = \sum_i (e_i e_i^\top) \sum_j \sigma^2_{ij} x_j \conj{x}_j
    \,.
\end{align*}
In a similar fashion we obtain the expression for the relation matrix of
the transformation:
\begin{equation*}
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    = \sum_i e_i e_i^\top \sum_j C_{ij} x_j^2
    \,.
\end{equation*}
The term $
  \sum_i e_i e_i^\top (\ldots)
$ means that the covariance and relation matrices of $y = Wx + b$ are diagonal.

These observations coupled with the closure under affine transformation imply that the
components of $y = W x + b$ are independent univariate $\cplx$-Gaussain random variables
with distributions
\begin{equation}  \label{eq:c-gaus-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \bigl(
        e_i^\top \mu x + b,
        \, \sum_j \sigma^2_{ij} \lvert x_j \rvert^2,
        \, \sum_j C_{ij} x_j^2
      \bigr)
    \,.
\end{equation}

We can notice, that in \eqref{eq:c-gaus-trick} the mean, variance and relation parameters are,
essentially, matrix-vector operations: $\mu x + b$, $\Sigma \lvert x \rvert^2$ and $C x^2$
with the modulus and power applied elementwise. The local reparameterization trick works
similarly for $\cplx$-convolutions, provided we draw independent weights $W\sim q(W)$ for each
spatial patch in the input: convolve value $x$, amplitude $\lvert x \rvert^2$ and complex
square $x^2$ with the mean, variance, and relation kernels, respectively. The need for
independent resampling stems from kernel weights being shared between patches, which would
make the gradient estimator less efficient due to correlation from overlapping patches.
The argument from \cite{kingma_variational_2015} in favour drawing independent weight per
sample in the minibatch can be applied here to show that this is \verify{indeed the case}.
%
Weight sharing is pushed from weights of the layer to shared variational parameters $\theta$
in the approximation $q_\theta$.

>>>>

%notation
We first clarify some notation: $M^{\top}$ is the matrix transpose, $\conj{M}$ is elementwise
complex conjugation, and $M^{\hop} = (\conj{M})^\top$ denotes Hermitian conjugate. By $e_i$
we denote the $i$-th real unit vector of dimensionality \textbf{conforming} to the matrix-vector
expression it is used in. $\vec{M}$ denotes row-major flattening of $M$ into a vector, i.e.
in lexicographic order of its indices. $\otimes$ denotes the Kronecker product, for which we
note the following identities $
  \vec{P Q R} = (P \otimes R^\top) \vec{Q}
$, and $
  (P \otimes R) (C \otimes S) = P Q \otimes R S
$, \cite{petersen_matrix_2012}.
% It follows from the definition of the Kronecker product
% of $P$ and $R^\top$ and the row-major order vectorization.
Finally, $
  \diag{}
    \colon \mathbb{K}^n \to \mathbb{K}^{n\times n}
    \colon v \mapsto (\delta_{ij} v_i)_{ij}
$ is a diagonal matrix embedding, $\mathbb{K} = \real$ or $\cplx$.

@@

$, respectively.
% $\cplx$-Gaussians are closed under affine transformations: if $y = A z + b$
% for $A \in \cplx^{n \times m}$ and $b \in \cplx^{n}$, then $
%   y \sim \mathcal{N}_n^{\cplx}\bigl(
%       A\mu + b, A \Gamma A^{\hop}, A C A^\top
%     \bigr)
% $.
$\cplx$-Gaussians are closed under affine transformations:
for $A \in \cplx^{n \times m}$ and $b \in \cplx^{n}$
\begin{equation}  \label{eq:cn-affine}
  A z + b \sim \mathcal{N}_n^{\cplx}\bigl(

@@

The entropy of a $\cplx$-Gaussian distribution coincides with the entropy of

@@

$C = 0$. From \eqref{eq:c-gauss-entropy-derivation} the entropy of a proper $\cplx$-Gaussian is

@@

In $\cplx$-variational dropout we use mean-field proper $\cplx$-Gaussian variational
approximation. Specifically, parameters $W \in \cplx^{n\times m}$ of a dense layer follow
\begin{equation}  \label{eq:c-gauss-vi}
  % (we omit the subscript $m=1$ for brevity)
  W
    \sim q(W)
    = \prod_{ij} q(W_{ij})
  \,,\,
    q(W_{ij})
      = \mathcal{N}^{\cplx}(W_{ij} \mid \mu_{ij}, \sigma^2_{ij}, 0)
  \,,
\end{equation}
and we use analogous independent proper $\cplx$-Gaussian approximation for convolutional
kernels. As for bias terms, we propose to use their MAP estimates, i.e. assume an independent
fully factorized Dirac approximation.

@@

The local reparameterization trick provides both faster computations, \cite{wang_fast_2013},
and better statistical efficiency of the SGVB gradient estimator, \cite{kingma_variational_2015}.
We formulate $\cplx$ version of the local reparameterization trick using a dense linear
layer as an example, and afterwards outline the caveats for $\cplx$-convolutions.

A general mean field $\cplx$-Gaussian approximation \eqref{eq:c-gauss-vi} for $
  W \in \cplx^{n\times m}
$ is
\begin{equation}  \label{eq:c-gauss-vi-general}
  \vec{W}
    \sim \mathcal{N}^{\cplx}_{nm} \bigl(
      \vec{\mu}, \diag{\vec{\Sigma}}, \diag{\vec{C}}
    \bigr)
  \,,
\end{equation}
where mean, \textit{variance} and relation are given by matrices $
  \mu, \Sigma, C \in \cplx^{n\times m}
$ with $\Sigma_{ij} \geq 0$, and $
  \lvert C_{ij} \rvert^2 \leq \Sigma_{ij}
$. For any $x \in \cplx^m$ and $b\in \cplx^n$ $
  y = W x + b
    = (I_n \otimes x^\top) \vec{W} + b
$, whence the covariance and relation matrices of $y$ are
\begin{align}
  (I \otimes x^\top) \diag{\vec{\Sigma}} (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (I \otimes x^\top) e_i \otimes e_j
    %   \Sigma_{ij} (e_i \otimes e_j)^\top (I \otimes x^\top)^{\hop}
    % = \sum_{ij} (e_i \otimes x^\top e_j)
    %   \Sigma_{ij} (e_i \otimes x^\top e_j)^{\hop}
    % = \sum_{ij} (e_i \otimes x_j) \Sigma_{ij} (e_i^\top \otimes \conj{x}_j)
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m \Sigma_{ij} x_j \conj{x}_j
    \,,  \label{eq:cn-gauss-lrt-cov} \\
  (I \otimes x^\top) \diag{\vec{C}} (I \otimes x^\top)^\top
    % = \sum_{ij} (e_i \otimes x_j) C_{ij} (e_i \otimes x_j)^\top
    & = \sum_{i=1}^n (e_i e_i^\top) \sum_{j=1}^m C_{ij} x_j^2
    \,.  \label{eq:cn-gauss-lrt-rel}
\end{align}
Since \eqref{eq:cn-gauss-lrt-cov} and \eqref{eq:cn-gauss-lrt-rel} are diagnal, $(y_i)_{i=1}^n$
are independent whence \eqref{eq:cn-affine} implies
\begin{equation}  \label{eq:cplx-gauss-trick}
  y_i
    \sim \mathcal{N}^{\cplx}
      \bigl(
        e_i^\top \mu x + b,
        \, \sum_{j=1}^m \Sigma_{ij} \lvert x_j \rvert^2,
        \, \sum_{j=1}^m C_{ij} x_j^2
      \bigr)
    \,.
\end{equation}
For a proper approximation \eqref{eq:c-gauss-vi} the resulting relation coefficients are zero.

\eqref{eq:cplx-gauss-trick} essentially requires three matrix-vector operations: $\mu x + b$,
$\Sigma \lvert x \rvert^2$ and $C x^2$ with the modulus and power applied elementwise. Similar
to $\real$ case, variational dropout for $\cplx$ convolutions draws independent kernel $W$
for each spatial patch in the input, since a convolution is a matrix product of $W$, unrolled
into a toeplitz matrix, and flattened intput $\vec{x}$. Weight sharing is essentially pushed
to shared variational parameters $\theta$ in the approximation $q_\theta$. The rationale
is to reduce correlation in the gradient estimator stemming from kernel weights being
shared between overlapping patches, \cite{kingma_variational_2015,citation_needed}. Thus
\eqref{eq:cplx-gauss-trick} requires computing $\cplx$ convolutions of $x$, complex square
$x^2$, and amplitude $\lvert x \rvert^2$ with the mean, relation, and variance kernels,
respectively.
